### 2016-08-12 robust-software.tex
Reorganizing the rules.
1. Minor editing on some existing rules.
2. Merged two rules about depending on external programs.
3. Reordered: 1, 2, 10, 3, 7/6, 5, 4, 9, 8
4. Added a paragraph for a new rule about keeping old versions available - needs work.

\begin{quote}
\begin{center}\textbf{Abstract}\end{center}
\fixme{GW}{Write abstract}
\end{quote}

@@

Typically, this software is developed by one person and 
produces excellent results in their hands. But what happens after the
student leaves someone else from their group wants to use the program? Everyone with
a few years of experience feels a tremor of fear when told, ``Use
Graduated Student's code to analyze your
data''. Often, that software will be undocumented and work in unexpected
ways, if it works at all, relying on nonexistent paths or resources, tuned for
the dataset, or simply an older version than was used in the paper.
The poor student or staff member then has two choices: hack the existing code
to make it work for her, or start over.

The root cause of this problem is that most of the software researchers
produce isn't \emph{robust}. The difference between running and being
robust is the difference between ``works for me on my machine'' and
``works for other people on a cluster I've never used''. In particular,
robust software:

\begin{itemize}
\item
  Works for users other than the original author
\item
  Is kept under version control.
\item
  Can be installed on more than one computer or account with relative ease
\item
  Has well-defined input and output formats
\item
  Has documentation that describes what its dependencies are, how to
  install it, and what the options are.
\item
  Comes with enough tests to show that it actually runs.
\end{itemize}

These rules need not be applied to \emph{every} coding effort. The vast majority of
code produced in the marathon of a graduate thesis is used once to answer a
specific question related to a specific dataset. Scientific endeavours often
require speed and rapid iteration over a question, not thoughtful software
engineering and thorough documentation. A README is of no use if you are the
sole user of your code. As the saying goes, not everything worth doing is worth
doing right away. However, once your script is dusted off and run three or four
times for slightly different purposes, when it comes time to publish your
methods, to leave a project, or to pass it on to someone else, it may be time
to apply ten simple rules for robust software.

These simple rules are also necessary steps toward creating a reusable library that
can be shared through a site like CPAN or CRAN. They can apply to both
closed-source and open-source software.  They are generic and can be applied to all
languages, libraries, packages, documentation styles, and
operating systems.  Whether the aim is as simple as sharing the code
with collaborators or as complex as using the software in a production
analysis environment, increasing the robustness of your software
decreases headaches all around.

@@

Input files and their formats are included in most documentation, but
intermediate, auxilliary and log files are often missing. 
\emph{All} files should be listed in the README, even those considered self-explanatory. Log
files are often full of valuable information that can be
mined for the user's specific purpose. Sometimes your users will need to answer a question of
the format, ``Does Software tell you the percentage of reads trimmed to remove
adapter sequences?'' and they can check the README and confidently
say ``yes, it is in the log file''.

\textbf{Attributions and licensing:} Attributions are how you credit
your main contributors; licenses are how you want others to use and
credit your software. By putting it in the README, it is readable \emph{before}
the software is installed (or even downloaded).
Leave no
question in anyone's mind about whether your software can be used
commercially, how much modification is permitted, and how other software
needs to credit you. If your software is not open source, state that clearly.
Attributions can also contain a list of `expert' users
that can be contacted if new users have problems with the software. (for
better or for worse)

\begin{itemize}
\item
  \fixme{GW}{example from real software:\\ https://github.com/dib-lab/khmer/blob/master/README.rst}
\end{itemize}

@@

@@

@@

\section{Reuse software (without tears).}

The way that the second program is invoked can throw up errors.
The program may not be in the user's path, or it may be an
older or newer version and produce results different than expected.
Windows users will be frustrated if you invoke bash or sh explictly or
use shell-specific conventions like `*' expansion.
Even Linux-standard functions available vary slightly between
installs. For example: GNU \texttt{sort} is available on almost every
*nix distribution, but sorts differently depending on locale.

We fully support the reuse of software between projects and have some
suggestions to reduce the aforementioned pains. First, 
make sure that you really need the call-out. If you are
executing GNU sort instead of figuring out how to sort lists in Python,
it may not be worth the pain of integration.

Second, ensure the
appropriate software and version is available. Either allow the user to
configure the exact path to the package, distribute the program with the
dependent software, or download it during installation using a
dependency management system. Regardless, check whether the program is
executable and what version is running. Be sure to remind users in the
documentation what the compatible versions are.

Finally, to ensure support on as many different operating systems as
possible, use native functions for starting other processes, such as
Java's Runtime.exec call, Python's subprocess module, and Perl's system
command, and be sure to capture and report the standard error output of
the subprocess to facilitate debugging.

\section{Do not rely on the pre-installation of non-standard packages or libraries unless clearly stated in the documentaton.}

Every package someone has to install before being able to use yours is a
possible (some would say ``likely'') source of frustration for some
potential user. On the other hand, research software developers should
re-use existing software wherever possible. To strike a balance between
these two, developers should document \emph{all} of the packages that
theirs depends on, preferably in a machine-readable form. For example,
it is common for Python projects to include a file called
\texttt{requirements.txt} that lists the names of required libraries,
along with version ranges:

A common way to break this rule is to depend on scripts and tools that
are installed on the computers the original developer is using, but
which aren't otherwise packaged and available. In many cases, the author
of a package may not realize that some tool was built locally, and
doesn't exist elsewhere. At present, the only sure way to discover such
unknown dependencies is to install on a system administered by someone
else and see what breaks. In future, as use of lightweight
virtualization containers like Docker becomes more widespread, it may
become common to test installation on a virtual machine.

>>>>

\begin{quote}
\begin{center}\textbf{Abstract}\end{center}
Software produced for research---even published software---suffers from a number of common problems
that make it difficult or impossible to run outside the original institution
or even the primary developer's computer.
We present ten simple rules to make software robust enough to run anywhere
and delight your users and collaborators.
\end{quote}

@@

Typically, this software is developed by one person and 
produces excellent results in their hands. But what happens after the
student leaves someone else from their group wants to use the program? Everyone with
a few years of experience feels a tremor of fear when told to use
a graduated student's code to analyze their data.
Often, that software will be undocumented and work in unexpected
ways (if it works at all).  It will often rely on nonexistent paths or resources,
be tuned for a single dataset,
or simply be an older version than was used in published papers.
Its new user is often faced with two unpalatable options:
hack the existing code to make it work, or start over.

The root cause of this problem is that most research software
isn't \emph{robust}. The difference between running and being
robust is the difference between ``works for me on my machine'' and
``works for other people on a cluster I've never used''. In particular,
robust software:

\begin{itemize}
\item
  works for users other than the original author;
\item
  is kept under version control;
\item
  can be installed on more than one computer or account with relative ease;
\item
  has well-defined input and output formats;
\item
  has documentation that describes what its dependencies are, how to
  install it, and what the options are; and
\item
  comes with enough tests to show that it actually runs.
\end{itemize}

These rules need not be applied to \emph{every} coding effort:
as the saying goes, not everything worth doing is worth doing right (or right away).
Code  that is used once to answer a specific question related to a specific dataset
doesn't require comprehensive documentation,
and the only sensible way to test it may well be to run it on the dataset in question.
However, if a script is dusted off and run three or four
times for slightly different purposes,
crucial to a publication,
or being passed on to someone else,
it may be time to apply ten simple rules for robust software.

These simple rules are also necessary steps toward creating a reusable
library that can be shared through a site like CPAN or CRAN.  They are
generic and can be applied to all languages, libraries, packages,
documentation styles, and operating systems for both closed-source and
open-source sofwtare.  Whether the aim is as simple as sharing the
code with collaborators or as complex as using the software in a
production analysis environment, increasing the robustness of your
software decreases headaches all around.

@@

Input files and their formats are included in most documentation, but
intermediate, auxilliary and log files are often missing. 
\emph{All} files should be listed in the README, even those considered self-explanatory. Log
files are often full of valuable information that can be
mined for the user's specific purpose. If your users might need to know,
``Does this program report the percentage of reads trimmed to remove
adapter sequences?'' they should be able to check the README and confidently
say, ``Yes, it is in the log file''.

\textbf{Attributions and licensing:} Attributions are how you credit
your main contributors; licenses are how you want others to use and
credit your software. By putting them in the README, it is readable \emph{before}
the software is installed (or even downloaded).
Leave no
question in anyone's mind about whether your software can be used
commercially, how much modification is permitted, and how other software
needs to credit you. If your software is not open source, state that clearly.
Attributions can also contain a list of ``expert'' users
that can be contacted if new users have problems with the software.

The README file for \withurl{khmer}{https://github.com/dib-lab/khmer/blob/master/README.rst}
is a good model:
it explains the software's purpose,
tells readers where to get help,
and includes links to a CITATION file (explaining how to cite the project)
and the license.

@@

\section{Make older versions available.}

Software evolves, but not all software evolves at the same pace.
While more consistent command-line arguments
or more flexible handling of output directories
may make a program better in general,
it can simultaneously make work for someone
who integrated the old version into their own workflow a year or two ago,
and won't see any benefits from upgrading.
Programs' authors should therefore ensure that every version they have released
continues to be available.

\fixme{MT}{Can you please flesh this out?}

@@

\section{Reuse software (within reason).}

In addition, every package someone has to install
before being able to use yours is a
possible (some would say ``likely'') source of frustration for some
potential user. On the other hand, research software developers should
re-use existing software wherever possible.

To strike a balance between
these two, developers should
\textbf{document \emph{all} of the packages that theirs depends on, preferably in a machine-readable form}.
For example, it is common for Python projects to include a file called
\texttt{requirements.txt} that lists the names of required libraries,
along with version ranges:

Conversely, developers should
\textbf{avoid depending on scripts and tools which are not available as packages},
even (or especially) if they are installed on the computers the original developer is using.
In many cases, a program's author may not realize that some tool was built locally, and
doesn't exist elsewhere. At present, the only sure way to discover such
unknown dependencies is to install on a system administered by someone
else and see what breaks. As use of lightweight
virtualization containers like Docker becomes more widespread, it may
become common to test installation on a virtual machine.

The way that the second program is invoked can throw up errors.
The program may not be in the user's path, or it may be an
older or newer version and produce results different than expected.
Windows users will be frustrated if you invoke \texttt{bash} or \texttt{sh} explictly or
use shell-specific conventions like `*' expansion.
Even Linux-standard functions available vary slightly between
installs. For example, GNU \texttt{sort} is available on almost every
*nix distribution, but sorts differently depending on locale.

Despite these caveats,
we fully support the reuse of software between projects provided
three additional guidelines are adhered to. First, 
\textbf{make sure that you really need the auxiliary program}. If you are
executing GNU sort instead of figuring out how to sort lists in Python,
it may not be worth the pain of integration.

Second, \textbf{ensure the appropriate software and version is available}.
Either allow the user to
configure the exact path to the package, distribute the program with the
dependent software, or download it during installation using a
dependency management system. Regardless, check whether the program is
executable and what version is running. Be sure to remind users in the
documentation what the compatible versions are.

Finally, to ensure support on as many different operating systems as
possible, \textbf{use native functions for starting other processes}, such as
Java's \texttt{Runtime.exec} call, Python's subprocess module, and Perl's system
command, and be sure to capture and report the standard error output of
the subprocess to facilitate debugging.

@@
